On the analytic structure of the linear response susceptibility in open systems

### 2018-03-15 AnalyticStructure_main.tex
update

Linear response can be effectively applied to the analysis of the electronic excitations in molecular and condensed-matter systems. In this framework one consider a system 
described by the Hamiltonian $\hnot$ and assume that the ground-state density $\dmnot$ has been provided. Linear response formalism allows us to evaluate the modification 
of the expectation value of a generic observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed 
through the evaluation of the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where the operator $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
under inspection. The modification of the ground-state density induced by the perturbation is expressed by the \emph{response density operator} $\dm'(\omega)$ that, in the linear 
response regime, is a linear function of the perturbing field $\op\Phi(\omega)$. For this reason it is possible to define the \emph{susceptibility} superoperator, as follows:
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
Susceptibility describes an intrinsic feature of the unperturbed system and determines its response to \emph{any} perturbations, in the linear response regime.
%ndependent from the specific perturbation that acts on the system and codifies the fundamental aspects which determine the response of the system. % is the fundamental object at the basis of

@@

The response density operator is expressed as the first-order variation of ground state density $\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$,
%\be\label{rhodef1}
%\dmnot = \sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}  \;,
%\ee
where the set $\{\ket{\psi_p}\}$ denote the occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level.
Using this notation we denote the energy of the HOMO level as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$. 
In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are real functions, when project in the $\r$-representation.

By considering the linear order time evolution of the eigenstates of $\hnot$ in presence of the perturbation we can derive the equation of motion of the response density in the form of 
a quantum Liouville equation (for example see \cite{baroni2008}). This procedure suggests us to introduce an explicit representation of the response density in terms of a set of 
$\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, as follows: 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Fluctuation states are, by construction, orthogonal to the occupied eigenstates of $\hnot$. Moreover, an analysis of the Liouville equation for the response density written in this fashion
evidences that their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
%An analysis of the Liouville equation for the response density written in this fashion evidences that fluctuation states are orthogonal to the occupied eigenstates 
%of $\hnot$ and their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'(\omega))\ket{\psi_p} \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$ and the operator $\op V'(\omega)$ is a perturbing term due to the density-dependence 
of the unperturbed Hamiltonian. %$\hnot$. %the unperturbed Hamiltonian. 
%and describes the perturbation induced by the variation of $\hnot$ under the action of the perturbing field.
%The  r.h.s of \eqref{fluctuationStateEqMotion1} expresses the complete perturbing potential

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a 
\emph{localization argument} for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential 
$\op V$ term and solve \eqref{fluctuationStateEqMotion1} in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r}\op G_H(\omega)\ket{\r'} = \frac{1}{4\pi} \left\{ 
% \begin{matrix}
% \frac{e^{-2k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega \leq |\eps_p| \\ 
% \frac{e^{\ii k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega > |\eps_p|
% \end{matrix}
% \right. \;,
% \ee
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\omega) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{2k|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; k^2 = \omega  < 0\; \\ 
\frac{e^{2\ii k |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; k^2 = \omega > 0
\end{cases} \;,
\ee
and the $\r$-representation of the fluctuation states $f_p(\omega,\r)$ is expressed through a convolution integral between the kernel \eqref{HelmholtzKernelDef1} and
$\brket{\r'}{s(\omega)}$.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, 
which govern the localization properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $\op G_H(\omega)$ behaves as a spherical wave and its
convolution with the source term always give raises to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
the kernel contains an exponential damping factor and %the localization features of $f_p(\omega,\r)$ are determined by the spatial behavior of the source term.  
$f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.

@@

% The knowledge of the long range behavior of the fluctuation states allows us to perform an assessment on the reliability of the evaluation of the linear response functional 
% \eqref{LinearResponseFunctDef1}. To elaborate this statement we make usage of the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1}, 
% so that:
% The knowledge of the long range behavior of the fluctuation states allows us to assess the spatial localization properties of the basic objects relevant in the contest of the 
% linear response. In particular, making usage of the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1}, we can express the linear 
% response functional \eqref{LinearResponseFunctDef1} as follows: 
This analysis allows us to comprehend that the spatial localization features of the building blocks relevant for the linear response depend critically on $\omega$. 
Indeed, the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1} implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}(\omega) = 2\sum_p \bra{\psi_p}\op O\ket{f_p(\omega)} \;,
\ee
so we can identify two different regimes, depending on the value of $\omega$. In the first one, realized for $\omega<|\eps_h|$, all the terms contributing to \eqref{LinearResponseFunctDef2} 
are expressed in terms of genuinely localized wave functions, while in the second one, for $\omega>|\eps_h|$, fluctuation states start behaving as unbound wave functions. 
%and the scalar products appearing in \eqref{LinearResponseFunctDef2} are only regularized by the long range behavior of the $\ket{\psi_p}$.
This fact implies that a computational setup has different chances of providing a reliable assessment of the linear response functional \eqref{LinearResponseFunctDef2} depending
to the value of $\omega$. 
This observation has relevant consequence on the capability of a computational setup to provide a reliable assessment of the linear response functional....

\vspace{0.3cm}
\noindent{TODO: Numerical example (convergence below threshold of the Benzene spectrum computed in different computational setup).}

>>>>

Linear response can be effectively applied to the analysis of the electronic excitations in molecular and condensed-matter systems. In this framework one consider a system 
described by the Hamiltonian $\hnot$ and assume that the ground-state density $\dmnot$ has been provided. Linear response formalism allows us to evaluate the modification 
of the expectation value of a generic observable induced by a time-dependent perturbing field acting on the system. This quantity, written in the frequency domain, can be expressed 
through the evaluation of the \emph{linear response functional}:
\be\lb{LinearResponseFunctDef1}
\braket{\delta\op O}(\omega) = \trace{\dm'(\omega)\op O} \;,
\ee
where the operator $\op O$ represents the observable\footnote{For a sake of concreteness, we are limiting our considerations to observables that do not explicitly depend on time} 
under inspection. The modification of the ground-state density induced by the perturbation is expressed by the \emph{response density operator} $\dm'(\omega)$ that, in the linear 
response regime, is a linear function of the perturbing field $\op\Phi(\omega)$. For this reason it is possible to define the \emph{susceptibility} superoperator, as follows:
\be\lb{SusceptibilityDef1}
\dm'(\omega) = \sop{{\cal R}}(\omega) \op\Phi(\omega) \;. 
\ee
Susceptibility describes an intrinsic feature of the unperturbed system and determines its response to \emph{any} perturbations, in the linear response regime.

@@

The response density operator is expressed as the first-order variation of ground state density $\sum_{\{p\}} \ket{\psi_p} \bra{\psi_p}$, where the set $\{\ket{\psi_p}\}$ denote the 
occupied states, \emph{i.e} the subset of the bound eigenstates of $\hnot$ with energy $\eps_p$ lower than the Fermi level. Using this notation we denote the energy of the HOMO level 
as $\eps_h$ and identify the value of the first ionization potential with $|\eps_h|$. In what follows we will assume that the set of states $\{\ket{\psi_p}\}$ are real functions, when 
projected in the $\r$-representation.

By considering the linear order time evolution of the eigenstates of $\hnot$ in presence of the perturbation we can derive the equation of motion of the response density in the form of 
a quantum Liouville equation (for example see \cite{baroni2008}). This procedure suggests us to introduce an explicit representation of the response density in terms of a set of 
$\omega$-dependent \emph{fluctuation states} $\{\ket{f_p(\omega)}\}$, as follows: 
\be\lb{rhoPrimeFluctuationStateDef1}
\dm'(\omega) = \sum_p\left(\ketbra{\psi_p}{f_p(-\omega)} + \ketbra{f_p(\omega)}{\psi_p}\right) \;.
\ee
Fluctuation states are, by construction, orthogonal to the occupied eigenstates of $\hnot$. Moreover, an analysis of the Liouville equation for the response density written in this 
fashion evidences that their equations of motion are written in the form of a modified Sternheimer equation \cite{mahan1980}:
\be\lb{fluctuationStateEqMotion1}
\left[\omega - (\hnot-\eps_p)\right]\ket{f_p(\omega)} = \op Q_0(\op\Phi(\omega)+\op V'(\omega))\ket{\psi_p} \;,
\ee
where $\op Q_0=\identity-\dmnot$ is the projector in the subspace orthogonal to the occupied states of $\hnot$ and the operator $\op V'(\omega)$ is a perturbing term due to the 
density-dependence of the unperturbed Hamiltonian.

The interesting aspect of equations \eqref{fluctuationStateEqMotion1} in the perspective of the present analysis is that they allow us to present a \emph{localization argument} 
for the fluctuation states. To clarify this statement we split the unperturbed Hamiltonian as the sum of the kinetic $\op T$ and potential $\op V$ term and solve \eqref{fluctuationStateEqMotion1} 
in the form of a self-consistent inhomogeneous Helmholtz equation by writing:
\be
\ket{f_p(\omega)} = \op G_H(\omega+\epsilon_p)\ket{s(\omega)} \;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\omega) = (\omega-\op T)^{-1}$, % with $k^2_p(\omega) = (\omega+\eps_p)$,
and the \emph{source term} is given by:
\be
\ket{s(\omega)} =  \op V\ket{f_p(\omega)} + \op Q_0(\op V'(\omega) + \op \Phi(\omega))\ket{\psi_p} \;.\nn
\ee
For open boundary conditions in three dimensions, the kernel of $\op G_H(\omega)$ in the coordinate representation reads: 
% \be\lb{HelmholtzKernelDef1}
% \bra{\r}\op G_H(\omega)\ket{\r'} = \frac{1}{4\pi} \left\{ 
% \begin{matrix}
% \frac{e^{-2k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega \leq |\eps_p| \\ 
% \frac{e^{\ii k_p(\omega)|\r-\r'|}}{|\r-\r'|} \;,\; {\textrm for} \; \omega > |\eps_p|
% \end{matrix}
% \right. \;,
% \ee
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\omega) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{2k|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; k^2 = \omega  < 0\; \\ 
\frac{e^{2\ii k |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; k^2 = \omega > 0
\end{cases} \;,
\ee
and the $\r$-representation of the fluctuation states $f_p(\omega,\r)$ is expressed through a convolution integral between the kernel \eqref{HelmholtzKernelDef1} and
$\brket{\r'}{s(\omega)}$.
An inspection of the structure of \eqref{HelmholtzKernelDef1} evidences that, for each value of $p$, there exist a threshold value $\omega = |\eps_p|$, 
which govern the localization properties of the corresponding fluctuation state. Indeed, for $\omega > |\eps_p|$,  $\op G_H(\omega)$ behaves as a spherical wave and its
convolution with the source term always give raises to a $f_p(\omega,\r)$ delocalized in the whole accessible space. On the contrary, for values of $\omega$ below the threshold,
the kernel contains an exponential damping factor and $f_p(\omega,\r)$ exhibit a bound-state behavior in the long range.

@@

% The knowledge of the long range behavior of the fluctuation states allows us to perform an assessment on the reliability of the evaluation of the linear response functional 
% \eqref{LinearResponseFunctDef1}. To elaborate this statement we make usage of the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1}, 
% so that:
% The knowledge of the long range behavior of the fluctuation states allows us to assess the spatial localization properties of the basic objects relevant in the contest of the 
% linear response. In particular, making usage of the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1}, we can express the linear 
% response functional \eqref{LinearResponseFunctDef1} as follows: 
% This analysis indicates that the task of providing a reliable assessment of the linear response functional present a different level of complexity   .
% Using the representation of the response density \eqref{rhoPrimeFluctuationStateDef1}, we can express the linear response functional as
% \be\lb{LinearResponseFunctDef2}
% \braket{\delta\op O}(\omega) = 2\sum_p \bra{\psi_p}\op O\ket{f_p(\omega)} \;.
% \ee
This argument can be easily extended to the linear response functional and allows us to determine the spatial localization features of the building blocks relevant for the linear response.  
Indeed, the fluctuation states representation of the response density \eqref{rhoPrimeFluctuationStateDef1} implies that
\be\lb{LinearResponseFunctDef2}
\braket{\delta\op O}(\omega) = 2\sum_p \bra{\psi_p}\op O\ket{f_p(\omega)} \;.
\ee
An inspection of \eqref{LinearResponseFunctDef2} evidences that we can identify two different regimes, on the basis of the value of $\omega$. The first one, which we will refer to as the
\emph{below threshold regime}, is realized for $\omega<|\eps_h|$. In this case the linear response functional is expressed in terms of genuinely localized quantities\footnote{This consideration 
holds for the class of observables for which $\bra{\psi_p}\op O$ is a localized state  with bound-state behavior in the long range}, each of which possesses a physical cut-off radius. 
Due to this behavior, the computational effort needed to provide a reliable evaluation of \eqref{LinearResponseFunctDef2} is determined by the capability of the chosen computational setup to 
express bound-state like wave functions, in a analogous fashion of a ground state calculation.

On the contrary, the evaluation of the linear response functional can be much more demanding in the \emph{above threshold regime}, realized for $\omega>|\eps_h|$, in which fluctuation 
states start behaving as unbound wave functions. In this case a computational description of fluctuation states is only meaningful in weak sense, \emph{i.e} inside the scalar products of 
\eqref{LinearResponseFunctDef2}, where the bound state character of $\{\bra{\psi_p}\op O\}$ behave as a regulator. This observation indicates that, above threshold, a computational setup 
provides a reliable assessment of \eqref{LinearResponseFunctDef2} only if it is able to express an unbiased description of the fluctuation states in the support of $\{\bra{\psi_p}\op O\}$.

As an explicit example in support of these arguments we compare the absorption spectrum of a benzene molecule computed in three distinct real-space basis set, which differs for the 
dimension of the computational domain. Results evidences that the curves are almost coincident in the below threshold regime, whereas the effect of the computational setup are clearly
visible in the for values of $\omega$ that exceed the threshold value.

% This fact implies that a computational setup has different chances of providing a reliable assessment of the linear response functional \eqref{LinearResponseFunctDef2} depending
% to the value of $\omega$. 
% This observation has relevant consequence on the capability of a computational setup to provide a reliable assessment of the linear response functional....

\vspace{0.3cm}
\noindent{TODO: Numerical example (convergence below threshold of the Benzene spectrum computed in different computational setup).}
\vspace{0.3cm}

This analysis, carried out starting from the representation of the linear response functional in terms of fluctuation states, evidences some general properties of the linear response. 
In the next sections we will reformulate an analogous procedure directly looking at the susceptibility. This analysis lead us to the core of the paper, namely the description of the
analytical structure of the linear response susceptibility.
